---
title: "Final Project"
author: "Haley Brown, Sabine Scott, Suzie Stadelman, & Emily Wilke"
format: 
  html:
    toc: true
    toc-depth: 5
    toc-title: "Table of Contents"
    code-fold: true
    code-summary: "Show the code"
editor: visual
bibliography: references.bib
link-citations: true
execute: 
  eval: true
  echo: true
---

# Code

## 1. Load Packages

```{r}

library(Lahman)
library(nlme)
library(janitor)
library(tidyverse)
library(rio)
library(mdsr)
library(here)
library(ggplot2)
library(dplyr)
library(ggridges)

##All but nlme has been cited in the .bib document. If any more packages get used will need to add citation. -Suzie
```

## 2. Read Data

```{r}
data <- import(here("data", "finalproj.csv")) %>%
  as_tibble()
```

## 3. Data Cleaning

### a. Demographics/Characteristics

```{r}

data$gender <- as.factor(data$gender)
data$gender

data$age <- as.factor(data$age)
data$age

data <- data%>%
  mutate(id= row_number())

data <- data %>%
  mutate(yrs_combine = recode(
    yrs_teach,
    "0-1 years" = "0-5 years",
    "2-5 years" = "0-5 years",
    "6-10 years" = "6-10 years",
    "11-15 years" = "11+ years",
    "More than 15 years" = "11+ years",
    .default ="Unknown"  # Change as necessary for unmatched values
  ))

data
```

### b. Measure Cleaning

#### i. Emotional Regulation Questionnaire (ERQ)

```{r}
erq_data_recode <- data %>%
  mutate(across(starts_with("er"), ~ case_when(.== "Strongly disagree" ~ 1,
                                               .=="Disagree" ~ 2,
                                               .=="Somewhat disagree" ~ 3,
                                               .=="Neither agree nor disagree (neutral)" ~ 4,
                                               .=="Somewhat agree" ~ 5,
                                               .=="Agree" ~ 6,
                                               .=="Strongly agree" ~ 7,
                                               .== "" ~ NA_real_)))

pro_data_recode <- erq_data_recode %>%
  mutate(across(starts_with("pro"), ~ case_when(.=="Never" ~ 1,
                                               .=="Rarely" ~ 2,
                                               .=="Sometimes" ~ 3,
                                               .=="Often" ~ 4,
                                               .=="Very often" ~ 5,
                                               .== "" ~ NA_real_)))

final_data <- pro_data_recode %>%
  mutate(across(starts_with("mhl"), ~ case_when(.== "Strongly disagree" ~ 1,
                                               .=="Disagree" ~ 2,
                                               .=="Somewhat disagree" ~ 3,
                                               .=="Somewhat agree" ~ 4,
                                               .=="Agree" ~ 5,
                                               .=="Strongly agree" ~ 6,
                                               .== "" ~ NA_real_)))



final_data<- final_data %>%
  mutate(mean_mhl_id = ((mhl_1 + mhl_2 + mhl_3 + mhl_4 + mhl_5 + mhl_6 + mhl_7 + mhl_8 + mhl_9 + mhl_10 + mhl_11 + mhl_11 + mhl_12 + mhl_13 + mhl_14 + mhl_15)/15))

final_data
```

#### i. Professional Quality of Life Scale (ProQOL)

```{r}
#Step 1: Reversals

#items 1, 2, 15, 17, 29
# 1 ->5
# 2 ->4
# 4-> 2
# 5 ->1

final_data <- final_data %>%
  mutate(pro_1 = as.numeric(recode(
    pro_1,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_2 = as.numeric(recode(
    pro_2,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_15 = as.numeric(recode(
    pro_15,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_17 = as.numeric(recode(
    pro_17,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_29 = as.numeric(recode(
    pro_29,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

#Step 2 Compassion Satisfaction

#add scores on 3, 6, 12, 16, 18, 20, 22, 24, 27, 30

final_data <- final_data %>%
  mutate(sum_comp_satis = (pro_3 + pro_6 + pro_12 + pro_16 + pro_18 + pro_20 + pro_22 + pro_24 + pro_27 + pro_30))

# Step 3 Burnout 
#add scores on 1, 4, 8, 10, 15, 17, 19, 21, 26, 29 
final_data

final_data <- final_data %>%
  mutate(sum_burnout = pro_1 + pro_4 + pro_8 + pro_10 + pro_15 + pro_17 + pro_19 + pro_21 + pro_26 + pro_29)


# Step 4 Secondary Traumatic Stress
#add scores on 2, 5, 7, 9, 11, 13, 14, 23, 25, 28 

final_data <- final_data %>%
  mutate(sum_sectrauma = pro_2 + pro_5 + pro_7 + pro_9 + pro_11 + pro_13 + pro_14 + pro_23 + pro_25 + pro_28)

#step 5 frequency table and histogram for compassion

frequency_table <- final_data %>%
  count(sum_comp_satis) %>%
  arrange(desc(n))
frequency_table
print(frequency_table)


comp_hist <- ggplot(final_data, aes(x=sum_comp_satis)) +
  geom_histogram()

comp_hist

#step 6 frequency table and histogram for burnout

frequency_table_2 <- final_data %>%
  count(sum_burnout) %>%
  arrange(desc(n))
frequency_table_2
print(frequency_table_2)


burnout_hist <- ggplot(final_data, aes(x=sum_burnout)) +
  geom_histogram()

burnout_hist

#step 7 frequency table and histogram for trauma 

frequency_table_3 <- final_data %>%
  count(sum_sectrauma) %>%
  arrange(desc(n))
frequency_table_3
print(frequency_table_3)


trauma_hist <- ggplot(final_data, aes(x=sum_sectrauma)) +
  geom_histogram()

trauma_hist
```

# Plot 1

```{r, plot-1}

#Mental Health Efficacy by School Type
#Do teacher reports of mental health efficacy differ by school type (i.e., elementary, middle, high school)? 

#There are 4 answer options on this question (Elementary, K-8, Middle, and High School), plus an other option.  
table(final_data$school_level)


#Six people wrote in answers. 4 said K-12, 1 said 6-8 school, and 1 said middle and high school. We could create a new category for K-12. I would feel comfortable putting the 6-8 person in the middle school category, and not sure what we'd want to do for the middle and high school. For now, I'm not going to include the "Other" option in the plot. 
table(final_data$school_level_other)

##I would like to rearrange the y-axis to be in order of level: Elementary, K-8, Middle, then High School. If you know how to do that, let me know! -Suzie

##Also there is one N/A for the mean mhl score, wondering if I need to add a na.rm somewhere? Lastly, I am getting a warning and I'm guessing that's what it's from. 

plot1 <- final_data %>%
  filter(school_level != "Other (Please Specify)") %>%
  select(school_level, mean_mhl_id) %>%
  ggplot(aes(x = mean_mhl_id, y= school_level)) +
  geom_density_ridges(fill = "lightblue", alpha = .7) +
  labs(x= "Mental Health Efficacy Score",
       y= "School Level",
       title = "Figure 1",
       subtitle = "Teacher Mental Health Efficacy by School Level") +
  theme_get()
  
         
plot1

#final_data %>%
  #ggplot(aes(school_type, mean_mhl_id) +
#geom_*** +
#theme

final_data %>%
    mutate(school_level= fct_collapse(school_level, 
        "Multiple Levels" = c("Other (Please Specify)", "K-8"),
        "Elementary School" = c("Elementary School"),
        "Middle School" = c("Middle School"),
        "High School" = c("High School")
  )) %>%
  ggplot() +
  geom_violin(aes(y = mean_mhl_id, x = school_level)) +
  ggtitle("Teacher Mental Health Efficacy \n by School Level") +
  xlab("Years of Experience") +
  ylab("Mental Health Efficacy") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

# Plot 2

```{r}
#Research question 2: Is there a relationship between teacher years of experience and teacher reports of mental health efficacy?

final_data %>%
    mutate(yrs_combine = factor(yrs_combine, levels = c("0-5 years", "6-10 years", "11+ years"))) %>%
  ggplot() +
  geom_boxplot(aes(y = mean_mhl_id, x = yrs_combine)) +
  ggtitle("Teacher Mental Health Efficacy \n by Years of Teaching Experience") +
  xlab("Years of Experience") +
  ylab("Mental Health Efficacy") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
  


```

# Plot 3

```{r}
#Research question 3: is there a relationship between teacher mental health efficacy and burnout? X = mental health efficacy, Y = burnout, scatter plot + line

final_data %>%
  ggplot(aes(mean_mhl_id, sum_burnout)) +
  geom_point() +
  geom_smooth() +
  ggtitle("Teacher Burnout by Mental Health Efficacy") +
  xlab("Mental Health Efficacy") +
  ylab("Burnout")+
  theme(plot.title = element_text(hjust = 0.5))
```

Research Questions (as reference)

1.  Do teacher reports of mental health efficacy differ by school level (i.e., elementary, middle, high school)? 

2.  Is there a relationship between teacher years of experience and teacher reports of mental health efficacy?

3.  Is there a relationship between teacher mental health efficacy and burnout?

Test

# Writing

### Abstract

### Introduction

### Methods

### Results

### Discussion

#### Reference Notes

here is one reference [@r]

here is another one @here

cite ERQ

cite ProQOL

cite nlme package
