---
title: "Final Project"    
subtitle: "EDLD 651"
author: "Haley Brown, Sabine Scott, Suzie Stadelman, & Emily Wilke" 
date: "December 11, 2024"
abstract: "Suzie will put Emily's abstract here"
format:                                                             
  html:                                                               
    toc: true                                                           
    toc-depth: 5                                                        
    toc-title: "Table of Contents"
    execute:
      eval: true
      echo: true 
      warning: false 
      error: false                                    
code-fold: true                                                     
code-summary: "Show the code"                                       
editor: visual                                                      
bibliography: references.bib                                        
link-citations: true                                                
fig-width: 6.5
fig-height: 8
---

# Writing

### Introduction

Concerningly, both teachers and students are facing mental health challenges [@agyapong2022; @cdc2023]. In attempts to promote student well being, social emotional learning has gained traction [@schwartz2022]. While teachers are being tasked with taking on increased responsibility to promote student mental health, little research has examined how equipped teachers feel to do this. Existing research suggests that teacher training in social-emotional learning and mental health is limited [@schiepe2021].

With proper preparation and support, teachers have the potential to be impactful in preventative mental health support for students. For example, teachers already participate in screening for academic and behavioral challenges, especially with the expansion of multi-tiered systems of support (MTSS), and response to intervention frameworks. Thus, expanding these teacher-led screening procedures to include social-emotional well being could be beneficial. Moreover, schools serve as a centralized resource for connecting and referring families to other community resources. This study seeks to address gaps in the current research on teachers’ ability to support study social-emotional well being by assessing teachers’ self-efficacy in supporting students with mental health concerns.

### Research Questions

The current study aims to develop an understanding of teachers’ perceptions of their mental health self-efficacy (MHSE) and burnout. Specifically, we asked the following questions:

1.  Do teacher reports of mental health efficacy differ by school type (i.e., elementary, middle, high school)? 

2.  Is there a relationship between teacher years of experience and teacher reports of mental health efficacy?

3.  Is there a relationship between teacher mental health efficacy and burnout?

### Methods

#### Participants and Recruitment

Oregon K-12 teachers were recruited during Spring 2024 to complete an anonymous, online survey. The goal of this survey was to understand how Oregon teachers viewed their use of emotion regulation, well-being, and mental health self-efficacy in the classroom. Principals across Oregon were contacted and asked if they would be willing to share the survey with their teaching staff. The final sample consisted of 92 teachers. Teacher demographic data can be found in [Table 1: Demographic Data].

#### Measures

***School Mental Health Self-Efficacy Teacher Scale (SMH-SETS)***

The School Mental Health Self-Efficacy Teacher Scale (SMH-SETS) is a 15-item measure that assesses teachers’ confidence in their ability to recognize, support, and refer students when they experience mental health concerns in the classroom [@brann2021]. Participants reported scores from strongly disagree (1) to strongly agree (6). Previous studies demonstrated that the SMH-SETS has acceptable reliability (Cronbach’s alpha = 0.91) and internal consistency ( = 0.93) [@brann2021].

***Professional Quality of Life (ProQOL) Scale***

The Professional Quality of Life (ProQOL) Scale measures positive and negative effects of working in a helping profession [@stamm2009]. This is a 30-item Likert-type scale, and participants rate questions from never (1) to very often (5). Items 1, 4, 15, 16, and 29 were reverse scored [@stamm2009]. For this study, the burnout subscale was calculated with items 1, 4, 8, 10, 15, 17, 19, 21, 26, and 29. Per the scoring manual, the average score on the burnout scale is 50 (SD= 10, alpha scale reliability =0.75).

#### Data Analysis

*Research Question 1: Do teacher reports of mental health self-efficacy differ by school type (i.e., elementary, middle, high school)?*

We created a violin plot (see [Plot 1]) to visualize any differences of mental health self-efficacy by school type. Our null hypothesis was that in the population of Oregon K-12 teachers, there is no difference in reports of mental health self-efficacy between elementary, middle, and high school teachers. We tested our null hypothesis via an Analysis of Variance (ANOVA). We also were curious to understand the magnitude by which teachers in different school types differed from each other in their mental health self-efficacy. We conducted an Ordinary Least Squares (OLS) estimation for each school type, with elementary schools as the reference, to interpret the results.

*Research Question 2: Is there a relationship between teacher years of experience and teacher reports of mental health self-efficacy?*

We created a box plot (see [Plot 2]) to visualize any differences of teacher MHSE by years of teaching experience. We were curious to understand the magnitude by which teachers’ MHSE differed by their years of teaching experience. We conducted an OLS estimation for teachers by years of teaching experience, with 11+ years of experience as the reference, to interpret the results.

*Research Question 3: Is there a relationship between teacher mental health efficacy and burnout?*

We created a scatter plot (see [Plot 3]) to visualize any relationship between teacher MHSE and reported burnout. We were curious to understand whether there was a significant relationship between teacher MHSE and burnout, so we conducted an OLS estimation for teachers’ MHSE and burnout. We also included the following covariates into our model: school level, years of experience, current class (general or special education).

### Results

```{r include = FALSE}
elementarymean <- 4.96
middlemean <- 4.90
highmean <- 5.07
multiplemean <- 4.98

```

*Research Question 1: Do teacher reports of mental health self-efficacy differ by school level (i.e., elementary, middle, high school)?*

The violin plot shown in [Plot 1] demonstrates no significant differences in average or spread of teacher mental health self efficacy by school level. Further, LINK-model 1 shows that there are no significant associations between teaching at middle school, high school, or multiple levels, and rating of mental health efficacy (p = 0.22, 0.28, and 0.27, respectively). The mean mental health efficacy of elementary teachers is `r elementarymean`, for middle school teachers is `r middlemean`, for high school teachers is `r highmean`, and for teachers who teach multiple grades is `r multiplemean`

*Research Question 2: Is there a relationship between teacher years of experience and teacher reports of mental health self-efficacy?*

The box plot shows in [Plot 2] demonstrates no significant differences in average or spread of teacher mental health self efficacy by years of experience. There appears to be a trend of lower self efficacy in teachers who have 6 - 10 years of experience, but [Model 2] shows no significant associations by years of experience (p > 0.05 for all predictors)

*Research Question 3: Is there a relationship between teacher mental health efficacy and burnout?*

Upon visual analysis, [Plot 3] demonstrates no significant differences in teacher burnout by mental health efficacy. To further explore this potential association, [Model 3] included covariates of school level, years of experience, and whether the teacher teaches general or special education. After accounting for these variables, there were still no significant associations between mental health self efficacy and teacher burnout.


### Discussion

# Code

## Load Packages

```{r}

#|include: false

library(Lahman)
library(nlme)
library(janitor)
library(tidyverse)
library(rio)
library(mdsr)
library(here)
library(ggplot2)
library(dplyr)
library(ggridges)
library(modelsummary)
library(gt)
library(sandwich)
library(kableExtra)

##All but nlme has been cited in the .bib document. If any more packages get used will need to add citation. -Suzie
```

## Read Data

```{r}
data <- import(here("data", "finalproj.csv")) %>%
  as_tibble()
```

## Data Cleaning

### Demographics/Characteristics

```{r}

data$gender <- as.factor(data$gender)
#data$gender

data$age <- as.factor(data$age)
#data$age

data <- data%>%
  mutate(id= row_number())

data <- data %>%
  mutate(yrs_combine = recode(
    yrs_teach,
    "0-1 years" = "0-5 years",
    "2-5 years" = "0-5 years",
    "6-10 years" = "6-10 years",
    "11-15 years" = "11+ years",
    "More than 15 years" = "11+ years",
    .default ="Unknown"  # Change as necessary for unmatched values
  ))

```

### Measure Cleaning

#### Emotional Regulation Questionnaire (ERQ)

```{r}
erq_data_recode <- data %>%
  mutate(across(starts_with("er"), ~ case_when(.== "Strongly disagree" ~ 1,
                                               .=="Disagree" ~ 2,
                                               .=="Somewhat disagree" ~ 3,
                                               .=="Neither agree nor disagree (neutral)" ~ 4,
                                               .=="Somewhat agree" ~ 5,
                                               .=="Agree" ~ 6,
                                               .=="Strongly agree" ~ 7,
                                               .== "" ~ NA_real_)))

pro_data_recode <- erq_data_recode %>%
  mutate(across(starts_with("pro"), ~ case_when(.=="Never" ~ 1,
                                               .=="Rarely" ~ 2,
                                               .=="Sometimes" ~ 3,
                                               .=="Often" ~ 4,
                                               .=="Very often" ~ 5,
                                               .== "" ~ NA_real_)))

final_data <- pro_data_recode %>%
  mutate(across(starts_with("mhl"), ~ case_when(.== "Strongly disagree" ~ 1,
                                               .=="Disagree" ~ 2,
                                               .=="Somewhat disagree" ~ 3,
                                               .=="Somewhat agree" ~ 4,
                                               .=="Agree" ~ 5,
                                               .=="Strongly agree" ~ 6,
                                               .== "" ~ NA_real_)))



final_data<- final_data %>%
  mutate(mean_mhl_id = ((mhl_1 + mhl_2 + mhl_3 + mhl_4 + mhl_5 + mhl_6 + mhl_7 + mhl_8 + mhl_9 + mhl_10 + mhl_11 + mhl_11 + mhl_12 + mhl_13 + mhl_14 + mhl_15)/15))

```

#### Professional Quality of Life Scale (ProQOL)

```{r}
#Step 1: Reversals

#items 1, 2, 15, 17, 29
# 1 ->5
# 2 ->4
# 4-> 2
# 5 ->1

final_data <- final_data %>%
  mutate(pro_1 = as.numeric(recode(
    pro_1,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_2 = as.numeric(recode(
    pro_2,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_15 = as.numeric(recode(
    pro_15,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_17 = as.numeric(recode(
    pro_17,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_29 = as.numeric(recode(
    pro_29,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

#Step 2 Compassion Satisfaction

#add scores on 3, 6, 12, 16, 18, 20, 22, 24, 27, 30

final_data <- final_data %>%
  mutate(sum_comp_satis = (pro_3 + pro_6 + pro_12 + pro_16 + pro_18 + pro_20 + pro_22 + pro_24 + pro_27 + pro_30))

# Step 3 Burnout 
#add scores on 1, 4, 8, 10, 15, 17, 19, 21, 26, 29 

final_data <- final_data %>%
  mutate(sum_burnout = pro_1 + pro_4 + pro_8 + pro_10 + pro_15 + pro_17 + pro_19 + pro_21 + pro_26 + pro_29)


# Step 4 Secondary Traumatic Stress
#add scores on 2, 5, 7, 9, 11, 13, 14, 23, 25, 28 

final_data <- final_data %>%
  mutate(sum_sectrauma = pro_2 + pro_5 + pro_7 + pro_9 + pro_11 + pro_13 + pro_14 + pro_23 + pro_25 + pro_28)


```

## Data Inspection using Frequency Tables and Historgrams

```{r}

#Compassion: frequency table and histogram

final_data %>%
  count(sum_comp_satis) %>%
  arrange(desc(n)) %>%
  kable(caption = "Frequency Table of Compassion Satisfaction",
        col.names = c("Compassion Satisfaction Score", "Count"))


comp_hist <- ggplot(final_data, aes(x=sum_comp_satis)) +
  geom_histogram(color= "gold4", fill= "darkgoldenrod1") +
  labs(x = "Compassion Satisfaction Score",
       y = "Count",
       title = "Histogram of Compassion Satisfaction",
       subtitle= "From the ProQOL Measure")

comp_hist

#Burnout: frequency table and histogram

final_data %>%
  count(sum_burnout) %>%
  arrange(desc(n)) %>%
  kable(caption = "Frequency Table of Burnout",
        col.names = c("Burnout Score", "Count"))

burnout_hist <- ggplot(final_data, aes(x=sum_burnout)) +
  geom_histogram(color= "darkolivegreen", fill= "darkolivegreen1") +
  labs(x = "Burnout Score",
       y = "Count",
       title = "Histogram of Burnout",
       subtitle= "From the ProQOL Measure")

burnout_hist

#Sec Trauma: frequency table and histogram

final_data %>%
  count(sum_sectrauma) %>%
  arrange(desc(n)) %>%  
  kable(caption = "Frequency Table of Secondary Traumatic Stress",
        col.names = c("Secondary Traumatic Stress Score", "Count"))


trauma_hist <- ggplot(final_data, aes(x=sum_sectrauma)) +
  geom_histogram(color= "deeppink4", fill= "deeppink") +
  labs(x = "Secondary Traumatic Stress Score",
       y = "Count",
       title = "Histogram of Secondary Traumatic Stress",
       subtitle= "From the ProQOL Measure")

trauma_hist

RQ1means <- final_data %>%
  group_by(school_level) %>%
  summarise(mean_mh = mean(mean_mhl_id, na.rm = TRUE))

RQ1means

```

# Table 1: Demographic Data

```{r}

#| label: table-1

final_data_clean_factor <- final_data %>%
  select(age, race_eth, yrs_teach, current_class) %>%
    rename(
    "Age" = "age",
    "Race/Ethnicity" = "race_eth",
    "Years of Teaching Experience" = "yrs_teach",
    "Current Classroom Type" = "current_class"
  ) %>%
  mutate(across(everything(), as.character))

 
# Pivot longer and calculate counts and percentages
demo_tbl <- final_data_clean_factor %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "level"
  ) %>%
  group_by(variable, level) %>%
  summarise(
    count = n(),
    percentage = n() / nrow(final_data_clean_factor) * 100,
    .groups = "drop"
  )

# Format table with gt, grouping rows by "variable"
demo_tbl %>%
  gt(groupname_col = "variable") %>%
  cols_label(
    level = "",
    count = "Count",
    percentage = "Percent (%)"
  ) %>%
  fmt_number(
    columns = percentage,
    decimals = 2
  ) %>%
  tab_header(
    title = md("Table 1: Demographic Data")
  ) %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_title()) %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) %>%
     tab_style(
       style = cell_fill(color = "ivory"),
       locations = cells_body()
     )
 


```

# Plot 1

```{r, plot-1}

#Mental Health Efficacy by School Type
#Do teacher reports of mental health efficacy differ by school type (i.e., elementary, middle, high school)? 

final_data <- final_data %>%
    mutate(school_level= fct_collapse(school_level, 
        "Multiple Levels" = c("Other (Please Specify)", "K-8"),
        "Elementary School" = c("Elementary School"),
        "Middle School" = c("Middle School"),
        "High School" = c("High School"))) 

ggplotrq1 <- final_data %>%
  ggplot(aes(y = mean_mhl_id, x = school_level, fill = school_level)) +
  geom_violin() +
  stat_summary(fun="median",
               geom = "crossbar",
               width = 0.5,
               colour = "red") +
  ggtitle("Plot 1: Teacher Mental Health Self-Efficacy \n by School Level") +
  xlab("Years of Experience") +
  ylab("Mental Health Efficacy") +
  labs(color = "School Level") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

ggplotrq1


```

# Model 1

```{r}
final_data$school_level <- factor(final_data$school_level, levels = c("Elementary School", "Middle School", "High School", "Multiple Levels"))

fit1 <- lm(mean_mhl_id ~ school_level, data=final_data)
anova(fit1)

modelsummary(fit1, 
             stars=T,
             fmt=2,
             vcov = "robust",
             gof_omit = "Adj.|AIC|BIC|Log|RMSE|RSE|Std.Err",
             coef_rename = c("school_levelMiddle School" = "Middle School",
                             "school_levelHigh School" = "High School",
                             "school_levelMultiple Levels" = "Multiple Levels"),
title = "Model 1. Ordinary Least Squares estimates of the relationship between the school level and average MHSE for Oregon teachers",
notes=c("Cells report coefficients and heteroscedastic-robust standard errors in parentheses. Each observation is one district."))

```

# Plot 2

```{r}
#Research question 2: Is there a relationship between teacher years of experience and teacher reports of mental health efficacy?

  
ggplotrq2 <- final_data %>%
  mutate(yrs_combine = factor(yrs_combine, levels = c("0-5 years", "6-10 years", "11+ years"))) %>%
  ggplot() +
  geom_boxplot(aes(y = mean_mhl_id, x = yrs_combine)) +
  ggtitle("Plot 2: Teacher Mental Health Self-Efficacy \n by Years of Teaching Experience") +
  xlab("Years of Experience") +
  ylab("Mental Health Self-Efficacy") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

ggplotrq2
  

```
# Model 2
```{r}
#OLS 

final_data$yrs_combine <- factor(final_data$yrs_combine, levels = c("11+ years", "0-5 years", "6-10 years"))

fit2 <- lm(mean_mhl_id ~ yrs_combine, data=final_data)

modelsummary(fit2, 
             stars=T,
             fmt=2,
             vcov = "robust",
             gof_omit = "Adj.|AIC|BIC|Log|RMSE|RSE|Std.Err",
             coef_rename = c("yrs_combine0-5 years" = "0-5 years",
                             "yrs_combine6-10 years" = "6-10 years"),
title = "Table ##. Ordinary Least Squares estimates of the relationship between the years of experience and average MHSE for Oregon teachers",
notes=c("Cells report coefficients and heteroscedastic-robust standard errors in parentheses. Each observation is one teacher"))


```

# Plot 3

```{r}
#Research question 3: is there a relationship between teacher mental health efficacy and burnout? X = mental health efficacy, Y = burnout, scatter plot + line

ggplotrq3 <- final_data %>%
  ggplot(aes(mean_mhl_id, sum_burnout)) +
  geom_point() +
  geom_smooth() +
  ggtitle("Plot 3: Teacher Burnout by Mental Health Self-Efficacy") +
  xlab("Mental Health Self-Efficacy") +
  ylab("Burnout")+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_get()

ggplotrq3


```

# Model 3
```{r}
#OLS 

fit3 <- lm(sum_burnout ~ mean_mhl_id + school_level + yrs_combine + current_class, data=final_data)

modelsummary(fit3, 
             stars=T,
             fmt=2,
             vcov = "robust",
             gof_omit = "Adj.|AIC|BIC|Log|RMSE|RSE|Std.Err",
title = "Table ##. Ordinary Least Squares estimates of the relationship between teacher MHSE and burnout for Oregon teachers",
notes=c("Cells report coefficients and heteroscedastic-robust standard errors in parentheses. Each observation is one teacher"))
```

