---
title: "Final Project"
author: "Haley Brown, Sabine Scott, Suzie Stadelman, & Emily Wilke"
format: 
  html:
    toc: true
    toc-depth: 5
    toc-title: "Table of Contents"
    code-fold: true
    code-summary: "Show the code"
editor: visual
bibliography: references.bib
link-citations: true
execute: 
  eval: true
  echo: true
---

# Code

## 1. Load Packages

```{r}

library(Lahman)
library(nlme)
library(janitor)
library(tidyverse)
library(rio)
library(mdsr)
library(here)
library(ggplot2)
library(dplyr)

##All but nlme has been cited in the .bib document. If any more packages get used will need to add citation. -Suzie
```

## Comment: I have learned about *Lahman* package from you, thanks.

## 2. Read Data

```{r}
data <- import(here("data", "finalproj.csv")) %>%
  as_tibble()
```

## 3. Data Cleaning

### a. Demographics/Characteristics

```{r}

data$gender <- as.factor(data$gender)
data$gender

data$age <- as.factor(data$age)
data$age

data <- data%>%
  mutate(id= row_number())

data <- data %>%
  mutate(yrs_combine = recode(
    yrs_teach,
    "0-1 years" = "0-5 years",
    "2-5 years" = "0-5 years",
    "6-10 years" = "6-10 years",
    "11-15 years" = "11+ years",
    "More than 15 years" = "11+ years",
    .default ="Unknown"  # Change as necessary for unmatched values
  ))

data
```

### b. Measure Cleaning

#### i. Emotional Regulation Questionnaire (ERQ)

```{r}
erq_data_recode <- data %>%
  mutate(across(starts_with("er"), ~ case_when(.== "Strongly disagree" ~ 1,
                                               .=="Disagree" ~ 2,
                                               .=="Somewhat disagree" ~ 3,
                                               .=="Neither agree nor disagree (neutral)" ~ 4,
                                               .=="Somewhat agree" ~ 5,
                                               .=="Agree" ~ 6,
                                               .=="Strongly agree" ~ 7,
                                               .== "" ~ NA_real_)))

pro_data_recode <- erq_data_recode %>%
  mutate(across(starts_with("pro"), ~ case_when(.=="Never" ~ 1,
                                               .=="Rarely" ~ 2,
                                               .=="Sometimes" ~ 3,
                                               .=="Often" ~ 4,
                                               .=="Very often" ~ 5,
                                               .== "" ~ NA_real_)))

final_data <- pro_data_recode %>%
  mutate(across(starts_with("mhl"), ~ case_when(.== "Strongly disagree" ~ 1,
                                               .=="Disagree" ~ 2,
                                               .=="Somewhat disagree" ~ 3,
                                               .=="Somewhat agree" ~ 4,
                                               .=="Agree" ~ 5,
                                               .=="Strongly agree" ~ 6,
                                               .== "" ~ NA_real_)))



final_data<- final_data %>%
  mutate(mean_mhl_id = ((mhl_1 + mhl_2 + mhl_3 + mhl_4 + mhl_5 + mhl_6 + mhl_7 + mhl_8 + mhl_9 + mhl_10 + mhl_11 + mhl_11 + mhl_12 + mhl_13 + mhl_14 + mhl_15)/15))

final_data
```

#### i. Professional Quality of Life Scale (ProQOL)

```{r}
#Step 1: Reversals

#items 1, 2, 15, 17, 29
# 1 ->5
# 2 ->4
# 4-> 2
# 5 ->1

final_data <- final_data %>%
  mutate(pro_1 = as.numeric(recode(
    pro_1,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_2 = as.numeric(recode(
    pro_2,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_15 = as.numeric(recode(
    pro_15,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_17 = as.numeric(recode(
    pro_17,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

final_data <- final_data %>%
  mutate(pro_29 = as.numeric(recode(
    pro_29,
    `1` = "5",
    `2` = "4",
    `3` = "3",
    `4` = "2",
    `5` = "1")))

### Comment: I love the way you coded reveresely. But have you thought about encapsulating the reversal logic in a reusable function and apply it dynamically:

# Function to reverse scores
reverse_scores <- function(data, vars) {
  data %>%
    mutate(across(all_of(vars), ~ as.numeric(recode(.,
      `1` = "5", `2` = "4", `3` = "3", `4` = "2", `5` = "1"
    ))))
}

# Reverse specific items
final_data <- reverse_scores(final_data, c("pro_1", "pro_2", "pro_15", "pro_17", "pro_29"))

#Step 2 Compassion Satisfaction

#add scores on 3, 6, 12, 16, 18, 20, 22, 24, 27, 30

final_data <- final_data %>%
  mutate(sum_comp_satis = (pro_3 + pro_6 + pro_12 + pro_16 + pro_18 + pro_20 + pro_22 + pro_24 + pro_27 + pro_30))

# Step 3 Burnout 
#add scores on 1, 4, 8, 10, 15, 17, 19, 21, 26, 29 
final_data

final_data <- final_data %>%
  mutate(sum_burnout = pro_1 + pro_4 + pro_8 + pro_10 + pro_15 + pro_17 + pro_19 + pro_21 + pro_26 + pro_29)


# Step 4 Secondary Traumatic Stress
#add scores on 2, 5, 7, 9, 11, 13, 14, 23, 25, 28 

final_data <- final_data %>%
  mutate(sum_sectrauma = pro_2 + pro_5 + pro_7 + pro_9 + pro_11 + pro_13 + pro_14 + pro_23 + pro_25 + pro_28)

###here as well. Good job on the summary. How about combining them together?
# Summation for scales
final_data <- final_data %>%
  mutate(
    sum_comp_satis = rowSums(across(c("pro_3", "pro_6", "pro_12", "pro_16", "pro_18", "pro_20", "pro_22", "pro_24", "pro_27", "pro_30")), na.rm = TRUE),
    sum_burnout = rowSums(across(c("pro_1", "pro_4", "pro_8", "pro_10", "pro_15", "pro_17", "pro_19", "pro_21", "pro_26", "pro_29")), na.rm = TRUE),
    sum_sectrauma = rowSums(across(c("pro_2", "pro_5", "pro_7", "pro_9", "pro_11", "pro_13", "pro_14", "pro_23", "pro_25", "pro_28")), na.rm = TRUE)
  )

#step 5 frequency table and histogram for compassion

frequency_table <- final_data %>%
  count(sum_comp_satis) %>%
  arrange(desc(n))
frequency_table
print(frequency_table)


comp_hist <- ggplot(final_data, aes(x=sum_comp_satis)) +
  geom_histogram()

comp_hist

#step 6 frequency table and histogram for burnout

frequency_table_2 <- final_data %>%
  count(sum_burnout) %>%
  arrange(desc(n))
frequency_table_2
print(frequency_table_2)


burnout_hist <- ggplot(final_data, aes(x=sum_burnout)) +
  geom_histogram()

burnout_hist

#step 7 frequency table and histogram for trauma 

frequency_table_3 <- final_data %>%
  count(sum_sectrauma) %>%
  arrange(desc(n))
frequency_table_3
print(frequency_table_3)


trauma_hist <- ggplot(final_data, aes(x=sum_sectrauma)) +
  geom_histogram()

trauma_hist
```

# Plot 1

```{r, plot-1}

#Mental Health Efficacy by School Type

table(final_data$school_type)
#There are 5 answer options here. 4 people selected two options. Need to decide how to clean this. 

#@suzie - for "Public school,Charter school" answers, let's rename this to be "Charter school" to fit with the other Charter school answers 

#@suzie let's keep dual language as public (so remane "Public school,Other" to "Public School")

table(final_data$school_type_other)
#Two people wrote in answers. Again need to decide how to clean this. 

# let's rename "Choice School" to "Public School" - we will make a note in the paper that we #had a dual language and choice school 

#@suzie, mean_mhl_id is the variable for SMH-SETS

#final_data %>%
  #ggplot(aes(school_type, mean_mhl_id) +
#geom_*** +
#theme


```

# Plot 3

```{r}
#Research question 3: is there a relationship between teacher mental health efficacy and burnout? X = mental health efficacy, Y = burnout, scatter plot + line

final_data %>%
  ggplot(aes(mean_mhl_id, sum_burnout)) +
  geom_point() +
  geom_smooth() +
  ggtitle("Teacher Burnout by Mental Health Efficacy") +
  xlab("Mental Health Efficacy") +
  ylab("Burnout")+
  theme(plot.title = element_text(hjust = 0.5))


### Comment: I love the way you guys adjust the element text. I learned a new thing about ggplot! I have tried the next plot code, hope you like it. 

final_data %>%
  ggplot(aes(x = mean_mhl_id, y = sum_burnout)) +
  geom_point(alpha = 0.7, color = "blue", size = 2) +  # Add transparency and color
  geom_smooth(method = "lm", se = TRUE, color = "red") +  # Explicit method for trend line
  ggtitle("Teacher Burnout by Mental Health Efficacy") +
  xlab("Mental Health Efficacy Score") +
  ylab("Burnout Level") +
  theme_minimal() +  # Use a clean theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),  # Enhance title
    axis.title = element_text(size = 12),  # Customize axis labels
    axis.text = element_text(size = 10)  # Adjust axis tick text size
  )


```

Research Questions (as reference)

1.  Do teacher reports of mental health efficacy differ by school type (i.e., elementary, middle, high school)? 

2.  Is there a relationship between teacher years of experience and teacher reports of mental health efficacy?

3.  Is there a relationship between teacher mental health efficacy and burnout?

Test

# Writing

### Abstract

### Introduction

### Methods

### Results

### Discussion

#### Reference Notes

here is one reference [@r]

here is another one @here

cite ERQ

cite ProQOL

cite nlme package
